---
title: "Car Thefts"
---

Follow along with the demo to learn how to calculate standard deviation and how to create a histogram of the data using R and the Tidyverse. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```



### Task 2: Load the car thefts dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv
```{r}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv")
```


### Task 3: Calculate the standard deviation of the number of car thefts, along with the mean

```{r}
car_thefts |> 
  summarise(mean_car_thefts = mean(count), sd_car_thefts = sd(count))


```



### Task 4: Create a histogram of the number of car thefts
### Task 5: Add a line to the histogram that shows the mean of the number of car thefts

```{r}
car_thefts |> 
  ggplot() + 
  geom_histogram(aes(x = count), binwidth = 2) +
  geom_vline(aes(xintercept = mean(count)), color = 'red', linetype = "dashed", size = 1) +
   geom_vline(aes(xintercept = mean(count) - sd(count)), color = 'blue', linetype = "dashed", size = 1) +
   geom_vline(aes(xintercept = mean(count) + sd(count)), color = 'green', linetype = "dashed", size = 1)

```




### Task 6: Describe your results

Is the distribution normal or skewed in one direction? Write a sentence in which you describe the range of the data covered by one standard deviation from the mean on both sides.

The majority of the data falls one standard deviation from the mean on either side. This data follows a semi-normal bell curve shape, with a slight skew left. There is significantly more data from the mean to one sd to the right. 


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
