---
title: "Crime Rates"
---

Follow along with the demo to learn how to calculate crime rates using R and the Tidyverse. We'll also introduce a new library, janitor, which helps make column names easier to work with. You should get started by loading the tidyverse in a new R block, then install and load janitor.

### Task 1: Load the tidyverse and install and load janitor
```{r}
library(tidyverse)
library(janitor)
```



### Task 2: Load the crime dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv
```{r}
crime_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv")
```



### Task 3: Use janitor to clean up the column names
```{r}
crime_data <- clean_names(crime_data)
```




### Task 4: Calculate the total crime rate for each year (picking our unit!)
```{r}
crime_data |>
  mutate(total_crime_rate = grand_total/population * 10000)
```



### Task 5: Calculate the violent and property crime rates for each year (same unit)
```{r}
crime_data |> mutate(violent_crime_rate = violent_crime_total/population * 10000)
crime_data |> mutate(property_crime_rate = property_crime_total/population * 10000)
```



### Task 6: Describe our results

Write a few sentences about what you found in the data. What years had the highest crime rates? What years had the lowest crime rates? Does the pattern match your expectations?

The earliest recorded years(1975-1983) have the highest crimes rates, until the mid 90s, when it spikes again. After 1997, it seems that the crime rate has been on a steady downward trajectory, where 2020, the most recent year in the data, had the lowest recorded crime rates. This matched my expectations for sure, I 've thought crime would be much lower now than it was in the 1970s and 1990s. 





When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

