---
title: "Volleyball Percentiles"
---

We'll be calculating percentiles to see how good Maryland was in service aces in the 2024 season. We'll use `mutate()` and a new function `ntile()` to calculate percentiles. Get started by loading the tidyverse in a code block, then load the CSV file here: https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/refs/heads/main/data/ncaa_womens_volleyball_matchstats_2024.csv" and save it to a dataframe called `matches`.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```


### Task 2: Load the data
```{r}
matches <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/refs/heads/main/data/ncaa_womens_volleyball_matchstats_2024.csv")

view(matches)
```


### Task 3: Create totals of aces for each team using `group_by()` and `summarize()`
```{r}
matches |> 
  group_by(team) |> 
summarise(total_aces=sum(aces)) |>
  arrange(desc(total_aces))
```


### Task 4: Calculate percentiles using `mutate()` and `ntile()`
```{r}
team_aces |> 
  mutate(percentile = ntile(total_aces, 100))

```



### Task 5: Describe the percentiles

Write a couple of sentences that puts Maryland into context with the other teams in the dataset. You can do some Internet research to help with this.

University of Maryland volleyball may have had a rough season, but when it comes to serving, they were able to hang with the best of the best. With 223 total aces, UMD volleyball ranks in the 95% percentile. While winning programs like UTEP and Sac St. lead the county, aces were a bright spot during a rough season for the terps. 




When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.






