---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)

```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv") 
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

ANSWER HERE

When looking through the data for the first time I immediately focus on the category that seems to have the most amount of incidents, which is accident. I am now assuming this means car accidents, however my first couple times looking over the data I wasn't exactly sure what that category was. Something that makes summarizing the data a bit harder is the three types of Breaking and Entering (B & E  COMMERCIAL, B & E  OTHER and B & E  RESIDENTIAL). Deciding to group these together or not definitely alters the data. Same thing with the different types of robbery. 




### Task 4: Calculating Averages

Calculate the total number of incidents for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |> 
  group_by(category) |> 
  #really struggled to figure out how to get a sum of the rows, as the category column is all characters not dbls  nor could I find anything in the book. I finally found rowSums while looking through different resources. Let me know if this was not the intended way to do it! 
mutate(total_incidents = rowSums(across(Jan:Dec), na.rm = TRUE)) |>
mutate(average_incidents = total_incidents / 12) 
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

ANSWER HERE

ACCIDENT, AUTO  STOLEN, THEFT FROM AUTO and ACCIDENT WITH IMPOUND are the four highest occurring crimes per month, which displays an obvious trend of crime revolving around cars. Excluding theft, there are 250 more incidents of the lowest car related crime than the highest non car related crime. 

### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |> 
  mutate(june_change = (Jun - average_incidents)/ Jun * 100) |> 
    mutate(july = (Jul - average_incidents)/ Jul * 100) |> 
    mutate(august_change = (Aug - average_incidents)/ Aug * 100) 
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

ANSWER HERE
For the majority of the types of crimes, there seems to be very little correlation between an increase during the summer months, however two stuck out. Both residential robbery and homicide increased (compared to the mean) during the three summer months. Although this tell us that the theory that crime is more likely to be committed in the summer months is statistically wrong, an increase in a major crime, homicide, that draws a lot of attention, may make it seem like an increase of overall crime.

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE

The most likely crime to be committed in Prince George's County has to do with your car, and its not even close. 

According to 2024 Prince George's County crime records, the three most commonly occurring crimes involve automobiles. These crimes are (in order of most common) car accidents, stolen vehicle and theft from vehicle.  These crimes are not only the most frequent but also account for the majority of monthly incidents, far surpassing other crime categories.


### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
