---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
mm_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.
```{r}
sample_5 <- mm_data |> 
  sample_n(5)

sample_10 <- mm_data |> 
  sample_n(10)
```



### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.
```{r}
sample_5 |> 
   summarise(mean_sample = mean(red), sd_sample = sd(red))
```

```{r}
sample_10 |> 
  summarise(mean_sample = mean(red), sd_sample = sd(red))
```



### Task 5: Calculate the mean and standard deviation of the population, using the same color as Task 4

```{r}
mm_data |>
  summarise(mean_sample = mean(red), sd_sample = sd(red))
```


### Task 6: Compare the sample and population statistics

Compare the three sets of means and standard deviations. Are they similar?

the 10 number sample and the dataset's mean and standard deviations are quite similar, about 3 apart in both mean and sd. On the other hand, the 5 number sample is more than 10 away from the actual mean just under 10 from the standard deviation. 

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

It is pretty obvious that the bigger the sample size, the closer the mean and standard deviation will be to that of the actual population. This is super valuable information that shows us that the larger and more random a sample is, the more representative it should be of the whole population. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
