---
title: "finalproject"
output: html_document
---

```{r}
install.packages('tidycensus')
library(tidycensus)
library(tidyverse)  # For data manipulation and visualization
library(lubridate)  # For handling dates
library(knitr)      # For nice tables
library(scales)     # For better axis formatting
library(sf)
library(tigris)

```



```{r}
#data load in
pg_crime_2023_2025 <- read_csv("~/Desktop/GitHub/Crime_Incidents_July_2023_to_Present_20250419.csv")
pg_crime_2023_2025_updated <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crimes_with_location.csv")
```
```{r}
#API Key
census_api_key("03183f1d7157fca8529a26898cff8c1b802565ed", overwrite = TRUE)
```

```{r}
view(pg_crime_2023_2025)
```

```{r}
#making a new data set with only specific sectors, and taking away accidents from the crime list. 
Grouped_data <- pg_crime_2023_2025_updated |>
   filter(
    !`PGPD Sector` %in% c("FXOS", "SHF", "P10", "FXIN"),
    !`clearance_code_inc_type` %in% c("ACCIDENT", "ACCIDENT WITH IMPOUND")
  ) |>
  group_by(zip_code) |>
  summarise(total_crimes = n(), .groups = "drop") |>
  arrange(desc(total_crimes))
```
```{r}
#i think this is the same, but don't want to delete it1 LOL
Grouped_by_zip <- pg_crime_2023_2025_updated |>
  filter(!`PGPD Sector` %in% c("FXOS", "SHF", "P10", "FXIN")) |>
  group_by(`PGPD Sector`, zip_code) |>
  summarise(total_crimes = n(), .groups = "drop") |>
  arrange(desc(total_crimes))

```

```{r}
#this code block displays the mean and standard deviation 
Grouped_data|> 
  summarise(mean_crimes = mean(total_crimes), sd_crimes = sd(total_crimes))
```
```{r}
#histogram of Crimes by Zip Code
ggplot(Grouped_data, aes(x = reorder(`zip_code`, -total_crimes), y = total_crimes)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = mean(Grouped_data$total_crimes), color = 'red', linetype = "dashed", size = 1) +
  geom_hline(yintercept = mean(Grouped_data$total_crimes) - sd(Grouped_data$total_crimes), color = 'blue', linetype = "dashed", size = 1) +
  geom_hline(yintercept = mean(Grouped_data$total_crimes) + sd(Grouped_data$total_crimes), color = 'green', linetype = "dashed", size = 1) +
  labs(x = "zip_code", y = "Total Crimes", title = "Total Crimes by Zipcode") +
  theme_minimal()
```
```{r}
Grouped_data


# Check the min max range in our data
min_crimes <- min(Grouped_data$total_crimes)
max_crimes <- max(Grouped_data$total_crimes)

```

```{r}
#creates z score 
Grouped_data <- Grouped_data |>
  mutate(
    mean_crime = mean(total_crimes),
    sd_crime = sd(total_crimes),
    z_score = (total_crimes - mean_crime) / sd_crime
  )
```
```{r}
# Display the data with z-scores
Grouped_data |>
  select('zip_code', total_crimes, z_score) |>
  head(10) 
```

```{r}
#makes category's labeled, based on Z score  
Grouped_data <- Grouped_data |>
  mutate(
    z_score = (total_crimes - mean(total_crimes)) / sd(total_crimes),
    z_category = factor(case_when(
      z_score >= 2.5 ~ "Very High",
      z_score >= .5 ~ "High",
      z_score >= -.5 ~ "Average",
      z_score >= -3 ~ "Low",
      TRUE ~ "Very Low"
    ), levels = c("Very High", "High", "Average", "Low", "Very Low"))
  )
```

```{r}
#making the histogram, same as the past one, this time color coded by High/Low Z_score 
ggplot(Grouped_data, aes(x = reorder(`zip_code`, -total_crimes), y = total_crimes, fill =  z_category)) +
  geom_col() +
  scale_fill_manual(values = c(
    "Very High" = "darkred",
    "High" = "orange",
    "Average" = "gray70",
    "Low" = "lightblue",
    "Very Low" = "blue"
  )) +
  labs(
    x = "zip_code",
    y = "Total Crimes",
    fill = "Z-Score Category",
    title = "Crime Totals by PGPD Sector by Z score"
  ) +
  theme_minimal()
```
#results show that 
```{r}
#Loading in census data 
acs_2022_5_year <- load_variables(2022, "acs5", cache = TRUE)

view(acs_2022_5_year)
```
#income by zip codes in PG County from 2022 census
```{r}
income_by_zip <- get_acs(
  geography = "zcta",
  variables = "B19013_001", 
  year = 2022,
  survey = "acs5" )
```


```{r}
#I Used AI for the following lines, To join the PG county zipcode  media income data from the census to my crime dada
zcta_shapes <- zctas(cb = TRUE, year = 2020)

# Get county boundaries for Maryland
md_counties <- counties(state = "MD", cb = TRUE)

# Transform projections to match
zcta_shapes <- st_transform(zcta_shapes, st_crs(md_counties))

# Spatial join to identify ZCTAs overlapping with Prince George's County
pg_zcta_shapes <- st_join(zcta_shapes, md_counties, join = st_intersects) |>
  filter(NAME == "Prince George's") |>
  select(GEOID20) %>%
  distinct()
```
```{r}
pg_income <- income_by_zip %>%
  filter(GEOID %in% pg_zcta_shapes$GEOID20)
```
```{r}
pg_county <- md_counties %>% filter(NAME == "Prince George's")

```



```{r}
pg_income_map <- pg_zcta_shapes %>%
  left_join(income_by_zip, by = c("GEOID20" = "GEOID"))

```


```{r}
#map in shape of PG, head map based on wealth by Zipcode
ggplot(data = pg_income_map) +
  geom_sf(aes(fill = estimate), color = "white") +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90", name = "Median Income") +
  labs(
    title = "Median Household Income by ZIP Code",
    subtitle = "Prince George's County, MD (ACS 2022)",
    caption = "Source: U.S. Census Bureau, ACS 5-Year Estimates"
  ) +
  theme_minimal()
```
```{r}
#joining the original data (crimes by zipcodes) with income by data
pg_crime_income <- Grouped_data |>
  left_join(pg_income, by = c("zip_code" = "GEOID"))
```
```{r}
zip_crime_counts <- pg_crime_income |>
  group_by(zip_code) |>
  summarise(total_crimes = n(), .groups = "drop")
```
```{r}
pg_crime_income_filtered <- pg_crime_income |>
  left_join(zip_crime_counts, by = "zip_code") |>
  filter(total_crimes.x > 500)
```
```{r}
pg_crime_income_under_500 <- pg_crime_income |>
  left_join(zip_crime_counts, by = "zip_code") |>
  filter(total_crimes.x < 500)
```

```{r}
pg_crime_income_non_filtered <- pg_crime_income |>
  left_join(zip_crime_counts, by = "zip_code") 
```



```{r}
#scatterplot to visualize the relationship between median income and crimes all zipcodes)
pg_crime_income_non_filtered |>
  ggplot(aes(x = `estimate`, y = `total_crimes.x`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "income", y = "total crimes", title = "Scatterplot of total_crimes vs median income")
```

```{r}
#scatterplot to visualize the relationship between median income and crimes (only zip codes with more than 500 total crimes)
pg_crime_income_filtered |>
  ggplot(aes(x = `estimate`, y = `total_crimes.x`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "income", y = "total crimes", title = "Scatterplot of total_crimes vs median income")
```
#this looks like there is a possible trend.
```{r}
#scatterplot to visualize the relationship between median income and crimes (only zip codes with less than 500 total crimes)

pg_crime_income_under_500 |>
  ggplot(aes(x = `estimate`, y = `total_crimes.x`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "income", y = "total crimes", title = "Scatterplot of total_crimes vs median income")
```
#looks like random points on a graph (seemingly no correlation )
```{r}
model <- lm(`total_crimes.x` ~ `estimate`, data = pg_crime_income_filtered) #(only zip codes with more than 500 total crimes)

summary(model)
```
#P Value higher than 0.05 but just barely. R Squared is around .15, maybe explains a bit.
```{r}
model_non <- lm(`total_crimes.x` ~ `estimate`, data = pg_crime_income_non_filtered) #(all zip codes )

summary(model_non)

```
#p-value way too high, not statisticly significnt. Adjusted R square is basiclly 0, meaning income not causing crimes. 
```{r}
pg_crime_income_filtered$predicted <- predict(model, newdata = pg_crime_income_filtered)

# Calculate the residuals
pg_crime_income_filtered <- pg_crime_income_filtered |> 
  mutate(residual = `total_crimes.x` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
pg_crime_income_filtered <- pg_crime_income_filtered |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))
```
#after messing around with the original dataset, the residuals are much closer than I originally calculated, however still are iffy at best. 

```{r}
pg_crime_income_non_filtered$predicted <- predict(model, newdata = pg_crime_income_non_filtered)

# Calculate the residuals
pg_crime_income_non_filtered <- pg_crime_income_non_filtered |> 
  mutate(residual = `total_crimes.x` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
pg_crime_income_non_filtered <- pg_crime_income_non_filtered |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))
```
#these are all over the place. The predicted is almost never close.

Headline:
“500 Crimes and Counting: Mapping the Invisible Line Between Income and Safety in Prince George’s County”

Main Findings:
Weak overall relationship between income and crime. When looking at all ZIP codes in Prince George’s County, the linear regression shows a very weak, statistically insignificant correlation between income and total crimes.R Squared = about 0.02  meaning income explains only about 2% of the variation in crime. p-value = 0.325 meaning not statistically significant.But once a ZIP code crosses the 500-crime mark, a pattern emerges
In ZIP codes with 500 or more crimes, the regression strengthens: R Squared = 0.15 meaning income now explains 15% of the variation.
Slope is negative (as expected): lower income correlates with more crime.
p-value = 0.0604 -> just above the typical 0.05 threshold, but strong enough to suggest a pattern worth investigating.
The trend indicates that ZIP codes with higher average incomes have notably fewer crimes, and the relationship becomes clearer once low-crime outliers are removed.
This suggests a tipping point effect:
Once crime passes a certain threshold (like 500), structural factors like poverty, segregation, and policing concentration begin to statistically track with income.
Potential Problems / Limitations:
Data Gaps: Several ZIP codes had missing income or crime totals and were excluded from the regression.
Causation vs. Correlation: Income alone doesn’t explain crime, other factors like policing policies, housing density, education access, and racial demographics are also at play.
ZIP Codes vs. Neighborhoods: ZIP codes are imperfect proxies for neighborhood-level dynamics; smaller geographies (like census tracts) could offer sharper insight.
Under reporting and Classification: Crimes like accidents and impounds were removed, but other categories may be inconsistently classified or reported.

Reporting Plan:

Source Interviews
Community Leaders in high-crime ZIP codes
PGPD officers or public information staff
Sociologists or urban economists who study poverty and crime

Field Reporting
Visit neighborhoods on either side of the 500-crime line to gather:
-Resident perspectives on safety
-Visible signs of policing, economic development, or abandonment
-Contrasts in housing, infrastructure, and retail activity

Narrative Angle
Follow the story of one ZIP code that defies the trend, a low-income area with relatively low crime, or a wealthy ZIP code with an unusually high crime rate to find into what other forces might be at play.

Why This Matters:
This story uses data to reveal a threshold effect in Prince George’s County: crime patterns aren’t random, and they don’t correlate with income everywhere, but once violence and theft rise above a certain point, structural poverty begins to take center stage. That’s a crucial insight for local policymakers, police reform advocates, and the communities themselves.
